Learning Discrete-valued Bayesian Network from Mixed Data

### 2015-10-26 BN_discretizer_paper.tex
update

The second solution is to discretize continuous variables. Discretization that learns from data has developed well and discussed \citep{Dougherty_1995} in the fields of machine learning and statistics for many year. Most of these discretization methods are designed for classification problems. They search the best discretization of a continuous attribute by considering its interation with the class variable of interest. However, these discretization methods do not apply to continuous variables in Bayesian network. In Bayesian network, interations and dependencies between variables are determined by graph structure. Therefore, if a discretization method only uses the interation between a continuous variable and its class variable of interest, instead of considering the graph structure, then it is not an appropriate discretization method for Bayesian networks. There are some reseach on discretizating continuous variables in naive Bayesian network and tree augmented network \citep{Fried_naive}. Nevertheless, only few discretization methods on general Bayesian netowrk have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998}.

The discretization technique proposed by N. Friedman and M. Goldszmidt \citet{Friedman_1996} is the most successful and practical one among these methods. It is based on minimum description length principle (MDL): optimal discretization policy minimizes the description length of Bayesian network and data information. If there is only one continuous variable in Bayesian network and other variables are discrete, MDL discretization method takes running time $O(N^3 + n_c {v_{max}}^{(n_c^p)_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, where $N$ is the number of data instances for learning the discretization, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_{max}$ is the largest cardinality number over all variables in Markov blanket, and  ${(n_c^p)_{max}}$ is the largest number of parent variables of the continuous variable's children.

In this paper, we propose a new discretization method for continuous variables in Bayesian network by learning from mixed data. This method is a generalized version of  \citet{Boulle_2006} and \citet{Lustgarten_2011}, which are both discretization methods of a continuous attribute with one class variable. We begin our method with the situation that only one variable in a given Bayesian network is continuous and other variable are discrete. Under this situation, we look for the most possible discretization policy $M$ given the data on other discrete variables. That is to say, the desired policy is $\textrm{arg}_M P(M|D)$. With Bayes rule, $P(M|D) = P(M) \cdot P(D|M) / P(D) \propto P(M)\cdot P(D|M)$. Usually, $P(D|M)$ increases as number of discretized intervals increases, since more intervals can provide more accuracy. Therefore, we design probability priors of $P(M)$ to decrease as number of intervals rises. As a result, we obtain a nature trade-off to determine the number of intervals after discretization. Besides, the $P(D|M)$ can be factorized according to Bayesian network structure. On the other hand, with the proposed priors, we are able to restrict the number of discretized intervals not exceeding the largest cardinality number of variable in Markov blanket too much. This is important, since for most algorithms on Bayesian network, their running times exponentially depend on the cardinality of variables. Another advantage of our algorithm is running time. If there is only one continuous variable, the running time is $O(n_c  {v_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, which beats the running time of MDL principle discretization \citep{Friedman_1996}. With an effective approximation proposed in this paper, we can further reduce running time to $O({v_{max}}(n_p+n_c)N^2)$.

@@

\section{Preliminaries}
\label{Prelim}
\subsection{Bayesian Network}
\label{BN_basic}
A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variable $X = \{  X_1, X_2, \cdots, X_n \}$, and whose edges $E$ represents probabilistic dependencies among nodes. The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents in $G$. The second component of $B$, namely $\Theta$, contains a set of parameter that qualifies the network. Elements of $\Theta$ take the form $\theta_{x_i|\Pi_{x_i}} = P(x_i|\Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$ (the set of parents of $X_i$ in $G$). Due to Markov property, we can represent the multivariate joint distribution over $X$ as
\begin{equation*}
P_B (X_1 , \cdots, X_n) = \prod_{i=1}^{n} P_B (X_i | \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i|\Pi_{x_i}}
\end{equation*}\\
For a given Bayesian network, we can generate its data instances by forward sampling, i.e., sampling variables one by one in a topological order \citep[see][chap.~22]{algo_2009}. Given a instance of $X_i$'s parent set $\Pi_{X_i}$, value of $X_i$ can be sampled according to the conditional probability table $P(X_i | \Pi_{X_i})$. Furthermore, this parent-child sampling order can be reversed if we know the marginal probability of child variable $P(X_i)$. By Bayes rule,
\begin{equation*}
P(X_i | \Pi_{X_i}) \cdot \prod_{j = 1}^{ Pa( X_i)} P( \{ \Pi_{X_i} \}_j) = P(X_i) \cdot P(\Pi_{X_i} | X_i),
\end{equation*}
where $Pa( X_i)$ is the number of parents of $X_i$, and $\{ \Pi_{X_i} \}_j$ is the $j$th parent of $X_i$. We can first sample $X_i$ by $P(X_i)$, then sample all the parent varaibles simultaneously by $P(\Pi_{X_i} | X_i)$. That is to say, $X_i$ becomes the starting point for sampling.

\subsection{K2 Structure Learning}
\label{intro_K2}
K2 ...

\section{Related Work}
\label{relat_work}
\subsection{MDL Principle Discretization}
\label{MDL}

>>>>

The second solution is to discretize continuous variables. Discretization that learns from data has developed well and discussed \citep{Dougherty_1995} in the fields of machine learning and statistics for many year. Most of these discretization methods are designed for classification problems. They search the best discretization of a continuous attribute by considering its interation with the class variable of interest. However, these discretization methods do not apply to continuous variables in Bayesian network. In Bayesian network, interations and dependencies between variables are determined by graph structure. Therefore, if a discretization method only uses the interation between a continuous variable and its class variable of interest, instead of considering the graph structure, then it is not an appropriate discretization method for Bayesian networks. There are some reseach on discretizating continuous variables in naive Bayesian network and tree augmented network \citep{Fried_naive}. Nevertheless, only few discretization methods on general Bayesian netowrk have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

The discretization technique proposed by N. Friedman and M. Goldszmidt \citet{Friedman_1996} is the most well-known one among these methods. It is based on minimum description length principle (MDL): optimal discretization policy minimizes the description length of Bayesian network and data information. If there is only one continuous variable in Bayesian network and other variables are discrete, MDL discretization method takes running time $O(N^3 + n_c {v_{max}}^{(n_c^p)_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, where $N$ is the number of data instances for learning the discretization, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_{max}$ is the largest cardinality number over all variables in Markov blanket, and  ${(n_c^p)_{max}}$ is the largest number of parent variables of the continuous variable's children.

In this paper, we propose a new discretization method for continuous variables in Bayesian network by learning from mixed data. This method is a generalized version of  \citet{Boulle_2006} and \citet{Lustgarten_2011}, which are both discretization methods of a continuous attribute with one class variable. We begin our method with the situation that only one variable in a given Bayesian network is continuous and other variable are discrete. Furthermore we assume that the network structure is known in advance. Under this situation, we look for the most possible discretization policy $M$ given the data on other discrete variables. That is to say, the desired policy is $\textrm{arg}_M P(M|D)$. With Bayes rule, $P(M|D) = P(M) \cdot P(D|M) / P(D) \propto P(M)\cdot P(D|M)$. Usually, $P(D|M)$ increases as number of discretized intervals increases, since more intervals can provide more accuracy. Therefore, we design probability priors of $P(M)$ to decrease as number of intervals rises. As a result, we obtain a nature trade-off to determine the number of intervals after discretization. Besides, the $P(D|M)$ can be factorized according to Bayesian network structure. On the other hand, with the proposed priors, we are able to restrict the number of discretized intervals not exceeding the largest cardinality number of variable in Markov blanket too much. This is important, since for most algorithms on Bayesian network, their running times exponentially depend on the cardinality of variables. Another advantage of our algorithm is running time. If there is only one continuous variable, the running time is $O(n_c  {v_{max}} \cdot N^2 + {v_{max}}^{n_p} \cdot N^2)$, which beats the running time of MDL principle discretization \citep{Friedman_1996}. With an effective approximation proposed in this paper, we can further reduce running time to $O({v_{max}}(n_p+n_c)N^2)$.

@@

\section{Preliminaries}
In this section we provide a brief review of Bayesian network, including the factorization of joint probability distribution, sampling from a given network, and strucutre learning. These concepts will be used in later sections.
\label{Prelim}
\subsection{Bayesian Network}
\label{BN_basic}
A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variable $X = \{  X_1, X_2, \cdots, X_n \}$, and whose edges $E$ represents probabilistic dependencies among nodes. The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents in $G$. The second component of $B$, namely $\Theta$, contains a set of parameter that qualifies the network. Elements of $\Theta$ take the form $\theta_{x_i|\Pi_{x_i}} = P(x_i|\Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$ (the set of parents of $X_i$ in $G$). Due to Markov property, we can represent the multivariate joint distribution over $X$ as
\begin{equation*}
P_B (X_1 , \cdots, X_n) = \prod_{i=1}^{n} P_B (X_i | \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i|\Pi_{x_i}}
\end{equation*}\\
For a given Bayesian network, we can generate its data instances by forward sampling, i.e., sampling variables one by one in a topological order \citep[see][chap.~22]{algo_2009}. Given a instance of $X_i$'s parent set $\Pi_{X_i}$, value of $X_i$ can be sampled according to the conditional probability table $P(X_i | \Pi_{X_i})$. Furthermore, this parent-child sampling order can be reversed if we know the marginal probability of child variable $P(X_i)$. By Bayes rule,
\begin{equation*}
P(X_i | \Pi_{X_i}) \cdot \prod_{j = 1}^{ Pa( X_i)} P( \{ \Pi_{X_i} \}_j) = P(X_i) \cdot P(\Pi_{X_i} | X_i),
\end{equation*}
where $Pa( X_i)$ is the number of parents of $X_i$, and $\{ \Pi_{X_i} \}_j$ is the $j$th parent of $X_i$. We can first sample $X_i$ by $P(X_i)$, then sample all the parent varaibles simultaneously by $P(\Pi_{X_i} | X_i)$. That is to say, $X_i$ becomes the starting point for sampling.

\subsection{K2 Structure Learning}
\label{intro_K2}
Roughly speaking, there are three approaches to learn a Bayesian network structure from data \citep[see][chap.~18]{PGM_2009}. They are constraint-based structure learning, score-based structure learning, and Bayesian model averaging. Here we review K2 structure learning algorithm \citep{K2}, which is 
one of the most successful score-based structure learning method. Same as most structure learning algorithms, K2 algorithm requires all variables are discrete. The score of a learned network is defined as $\prod_i f(X_i, \Pi_{X_i})$, where
\begin{align}
f(X_i, \Pi_{X_i}) = \prod_{j=1}^{|\Pi_{X_i}|} {{(|X_i|-1)!}\over{(N_{ij} + |X_i|-1)!}} \prod_{k=1}^{|X_i|} \alpha_{ijk}!.
\end{align}
$|X_i|$ is the cardinality of variable $X_i$. $|\Pi_{X_i}|$ is number of all possible instantiations of the parent variables of $X_i$, i.e., $|\Pi_{X_i}| = \prod_{Y \in \Pi_{X_i}} |Y|$. $\alpha_{ijk}$ is number of instances in data set that variable $X_i$ is instantiated with its $k$th value, and the parent of $X_i$ are instantiated with the $j$th value of $\Pi_{X_i}$. $N_{ij} = \sum_{k=1}^{|X_i|} \alpha_{ijk}$. K2 algorithm searchs the network structure with highest score, which can be interpreted as the most probable network with the given data. Besides, K2 requires the topological order of variables be known before scoring can start. This constraint can prevents cycles from being introduced. The searching for high-score network is a iterative process. There is no way to find the optimal network directly, since there are $2^{O(n^2)}$ possible structures, where $n$ is number of variables. In order to compensate for this, we run K2 algorithm for many times, and each time we start with a different order of variables. The network with highest score in iterations is the desired one.

\section{Related Work}
\label{relat_work}
In these section we review two related works: MDL principle discretization \citep{Friedman_1996} and MODL discretization \citep{Boulle_2006}. The former is the most famous discretization method for continuous variables in Bayesian network, and we will compare it with our proposed method in Section 6. The latter is a discretization method for one continuous attribute according to a target class. Our proposed method is a generalization of this method.   
\subsection{MDL Principle Discretization}
\label{MDL}
